---
title: "Male-Female fertility differences at the regional level"
author: "Henrik Schubert"
date: "2023-06-06"
output: pdf_document
bibliography: U:/male fertility/Male_Fertility_Mexico/Text/references.bib
abstract: "In this file, male and female fertility rates in Mexico at the regional level are compared. We will use graphical tools, distributional comparison measures, decomposition methods and statistical tests. The comparisons will be both over time, and at the national level "
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load the functions and packages
source("Functions/Packages.R")
source("Functions/Functions.R")
source("Functions/Graphics.R")

# Load the national level data
load("Data/tfr_regional_mexico.Rda")
```


```{r}
### Plot the distribution over time
tfr_reg %>% 
  pivot_longer(cols = starts_with("tfr"), names_prefix = "tfr_", values_to = "tfr", names_to = "sex") %>% 
  filter(sex != "TFR_ratio") %>% 
  ggplot(aes(as.factor(year), tfr, fill = sex, colour = sex)) +
  scale_fill_manual(values = c( MPIDRred, MPIDRblue), name = "Sex:") +
  scale_colour_manual(values = c( MPIDRred, MPIDRblue), name = "Sex:") +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = c(0.8, 0.9)) +
  ylab("Total fertility rate") +
  xlab("Year")
```


# Sub-national level
Now, we are looking at the variation at the sub-national level.

```{r}
# What is the trend in the national tfr s
ggplot(tfr_reg, aes(year)) +
  geom_line(aes(y = tfr_f, colour = "female", group = entity), linewidth = 1.4, alpha = 0.5) +
  geom_line(aes(y = tfr_m, colour = "male", group = entity), linewidth = 1.4, alpha = 0.5) +
  annotate(geom = "text",
           x = 2018, y = 4, 
           label = "Male", 
           colour = MPIDRblue, size = 8, family = "serif") +
  annotate(geom = "text",
           x = 1998, y = 1.8,
           label = "Female",
           colour = MPIDRred, size = 8, family = "serif") +
  scale_colour_manual(values = c(MPIDRred, MPIDRblue)) +
  scale_x_continuous(expand = c(0, 0)) +
  ylab("Total fertility rate") +
  xlab("Year") +
  guides(colour = "none")
```


```{r}
# Load the national level data
load("Data/asfr_regional_mexico.Rda")

# What is the trend in the national TFR for women
ggplot(asfr_reg, aes(x = age, alpha = year)) +
  geom_line(aes(y = asfr_f, colour = entity_name, group = interaction(year, entity)),
            linewidth = 1.4) +
  ylab("Age-specific fertilitly rate") +
  xlab("Age") +
  facet_wrap(~ entity_name) +
  guides(alpha = "none", colour = "none")
```


```{r}
# What is the trend in the national TFR for women
ggplot(asfr_reg, aes(x = age, alpha = year)) +
  geom_line(aes(y = asfr_m, colour = entity_name, group = interaction(year, entity)),
            linewidth = 1.4) +
  ylab("Age-specific fertilitly rate") +
  xlab("Age") +
  facet_wrap(~ entity_name) +
  guides(alpha = "none", colour = "none")
```

The graphs are overfull. In order to make the graphics comparable, we estimate the male-to-female asfr ratio

```{r}
# Plot the age-specific fertility rates
asfr_reg %>% 
  mutate(asfr_ratio = asfr_m / asfr_f) %>% 
  ggplot(aes(x = age, alpha = year)) +
  geom_line(aes(y = asfr_ratio, colour = entity_name, group = interaction(year, entity)),
            linewidth = 1.4) +
  geom_hline(yintercept = 1) +
  ylab("Age-specific fertilitly rate") +
  scale_y_continuous(trans = "log10", labels = scales::label_number_si()) +
  xlab("Age") +
  facet_wrap(~ entity_name) +
  guides(alpha = "none", colour = "none")
```




### Decomposition of regional differences


```{r}
# Estimate the decomposition
comp_asfr_reg <- asfr_reg %>%
                group_by(year, entity_name, entity) %>%
                mutate(pop_share_f = pop_share(mid_year_pop_f),
                       pop_share_m = pop_share(mid_year_pop_m),
                       delta_pop = difference(pop_share_f, pop_share_m),
                       delta_rate = difference(asfr_f, asfr_m),
                       change_rate = pop_share_f * delta_rate,
                       change_pop  = asfr_f * delta_pop,
                       change_inter = delta_pop * delta_rate)
  
# Make the decomposition
decomp_reg <- comp_asfr_reg %>%
  select(year, age, change_rate, 
         change_pop, change_inter, entity_name) %>%
  pivot_longer(cols = starts_with("change"),
               names_to = "component",
               values_to = "contribution") 

# Plot the decomposition
decomp_reg %>% 
  filter(year %in% c(1990, 2020) &
         entity_name %in% c("Baja California Sur",
                            "Guerrero")) %>% 
  ggplot(aes(age, contribution, fill = component)) +
  geom_col(colour = "white", linewidth = 0.01) +
  geom_hline(yintercept = 0, colour = "black") +
  facet_grid(entity_name ~ year) +
  scale_fill_manual(values = c(MPIDRred, MPIDRorange, MPIDRblue),
                    labels = c(expression(paste(Delta, "Population and rate")),
                             expression(paste(Delta, "Population")),
                             expression(paste(Delta, "Rate"))),
                    name = "Component: ") +
  scale_x_continuous(expand = c(0, 0)) +
  ylab("Contribution") + xlab("Age") +
  theme(legend.key.width = unit(0.3, "cm"),
        legend.key.height = unit(0.2, "cm"))
```

In order to better understand the size of the contribution, we create tables that display the components with the largest contribution.


```{r}
# Create a table
left_join(decomp_reg, tfr_reg, by = c("year", "entity")) %>%
  mutate(change = tfr_f - tfr_m,
         relative_contribution = contribution / change, 2) %>% 
  filter(year == 1990) %>% 
  ungroup() %>% 
  select(component, entity_name, age, contribution,
         relative_contribution, tfr_f, tfr_m, change) %>% 
  arrange(desc(abs(relative_contribution))) %>%
  mutate(relative_contribution =  paste(round(100 * relative_contribution, 2), "%")) %>%
  slice_head(n = 5) %>% 
  pander()
```